Untitled

### 2022-10-29 Draft_1.Rmd
Removing useless code chunks and adding in interpretation of chosen basic model

```{r,echo=F}
#Histogram to examine the expenditure
hist(expend.df$expenditure,breaks = 20,
     xlab = "Total weekly household expenditure (GBP)",
     main = "Histogram of expenditure")

@@

grid.arrange(p1,p2,p3,p4,p5,p6,ncol=2,nrow=3)
```
```{r}
p6
```

Looking at the above histogram of expenditure it shows that expenditure has a positively skewed distribution. Noticing this after applying a log transform the histogram of log expenditure has a more normal looking distribution.

Investigating covariates against expenditure we can see that there is a positive association between household income and expenditure.
For the factor variables I use box plots to examine any differences present between the different levels expenditure.
Household tenure displays differences that public rented households have noticeably lower expenditure compared to the other 2 household levels
Female household leads expend less on average
Employment status appears to be associated with expenditure with unemployed having a dramicatlly lower expenditure compared to full and part time employment categories
Household size appears to be positively correlated with expenditure, with larger households having higher expenditure except for 5+ person households which goes against this trend (further investigation?)
Number of adults in the household also appears positively correlated with households containing more adults having a higher expenditure

```{r,echo=F,eval=F}
#extra code that may not use
pairs(expend.df)
xtabs(~expend.df$house.ten+expend.df$sex.hh)

par(mfrow=c(1,2))
hist(expend.df$expenditure,breaks = 20,
     xlab = "Total weekly household expenditure (GBP)")
hist(log(expend.df$expenditure),breaks = 20,
     xlab = " Log of total weekly \n household expenditure (GBP)")
#Applying log transform gives a more normal looking distribution
```

@@

```{r,echo=F}
#Question 2
lm.income<-lm(expenditure~income,data = expend.df)
#summary(lm.income)
as.data.frame(tidy(lm.income)[,1:3])%>%
  kable(escape = F)%>%
  kable_styling(font_size = 15)
par(mfrow=c(2,2))
plot(lm.income)
# #as expected the normal QQ plot shows 'bad fit'

# #try gamma as good for postively skewed continous data
# gamma.income <-glm(expenditure~income,data=expend.df,family="Gamma")
# summary(gamma.income)
# plot(gamma.income)
# #bad fit (explain)
# 
# #out of all models log linear is the most appropriate
```
Fitting a linear model $y_{i}=\beta_{0}+\beta_{1}x_{i}+\epsilon$ with the response $y$ the expenditure,the explanatory variable $x$ the income and $i \in \{1,...,1200\}$. The above table shows the estimated values of the intercept ($\beta_0$) and the estimated 'effect' of a 1 pound increase in income on expenditure $\beta_1$ and there standard errors.

@@

#### Part 5:
```{r,echo=F}
lm.log.income.sqd <- lm(log(expenditure)~income+I(income^2),data = expend.df)
tidy(lm.log.income.sqd)[,1:3]%>%
  kable(escape = F)%>%
  kable_styling(font_size = 15)
par(mfrow=c(2,2))
plot(lm.log.income.sqd)
```
Fitting the model $log(y_{i})=\beta_{0}+\beta_{1}x_{i}+\beta_{2}x^2_{i}+\epsilon$ gives the estimated coefficients above. Viewing the diagnostic plots the fitted vs residuals plot shows a fairly even scattering of points around 0 thus the linearity assumption appears valid. The QQ plot shows the points stay close to the straight line meaning the normality assumption is valid.

@@

Using AIC as a measure of fit the best is model 3. However model 4 has the highest $R^2$ and adjusted $R^2$ suggestiong that model 4 is the best fit to the data. Which is further backed up by the diagnostic plots giving the indication that this is the only model in which the linearity and normality assumptions are met. Due to the diagnostic plots and $R^2$ values I choose model 4 as my preferred model.

```{r}
tidy(lm.log.income)
tidy(lm.log.income.sqd)
```

Using this model a person who has no income is expected to spend £4.71 per week (add in CI) and a £100 increase in income per week will increase expenditure on average by $exp(3.12x-0.00145x^2)$

MODEL IS LOG SO NEED TO EXP IT!!!!

#### Task 7
```{r}
#Investigating possible interactions
ggplot(data=expend.df,aes(x=sex.hh,fill=hh.adults,y=expenditure))+geom_boxplot()
```

```{r}
#Creating model with all possible 2 way interactions (not a good idea but just for interest :P)
lm.log.full<-lm(log(expenditure)~income+I(income^2)+income*sex.hh+income*hh.adults+income*house.ten+income*lab.force+income*hh.size+sex.hh*hh.adults+sex.hh*house.ten+sex.hh*lab.force+sex.hh*hh.size+hh.adults*house.ten+hh.adults*lab.force+hh.adults*hh.size+house.ten*lab.force+house.ten*hh.size+lab.force*hh.size,data=expend.df)
summary(lm.log.full)

step.method<- step(lm.log.full,trace=F)
summary(step.method)

```

```{r}
#Using AIC as the model fitting criteria (step function)
#Note that we are allowed to use interactions!
lm.log.full<-lm(log(expenditure)~income+I(income^2)+income*sex.hh+income*hh.adults+income*house.ten+lab.force+hh.size,data=expend.df)
summary(lm.log.full)
step.method<- step(lm.log.full,trace=F)
summary(step.method)
AIC(step.method)
plot(step.method)
```

@@

#### Model selection
For the model selection process I first wanted to decide which transformation is best to apply to the response as it is clearly skewed. using the box

For this model selection process. I first decided to only consider 2 way interactions as higher order interactions make interpretation of the model much more complicated.
When considering which interactions to include I first created plots to examine if there are any clear interactions between income and the other covariates. Considering the factor covarites thinking about the possible interactions none made much logical sense e.g. a household female head would have a different effect on expenditure to male depending on the number of people in the household. Thus didn't include any factor:factor ionteractions in the model.

```{r}
#interaction between sex and income
interaction.continous <-function(var){
  ggplot(data=expend.df,aes(x=income,y=expenditure,colour=.data[[var]]))+
    geom_point()
}

@@

```{r}
lm.log.full<-lm(log(expenditure)~income+I(income^2)+income*sex.hh+income*hh.adults+income*house.ten+lab.force*income+hh.size*income,data=expend.df)
summary(lm.log.full)
step.method<- step(lm.log.full,trace=F)
summary(step.method)
AIC(step.method)
plot(step.method)

par(mfrow=c(1,2))
hist(sqrt(expend.df$expenditure),breaks = 20)
hist(log(expend.df$expenditure),breaks = 20)

lm.sqrt<-lm(sqrt(expenditure)~income+I(income^2),data = expend.df)
plot(lm.sqrt)
```

>>>>

#### Part 1:
```{r,echo=F}
#Histogram to examine the expenditure
hist(expend.df$expenditure,breaks = 20,
     xlab = "Total weekly household expenditure (GBP)",
     main = "Histogram of expenditure")

@@

grid.arrange(p1,p2,p3,p4,p5,p6,ncol=2,nrow=3)
```

Looking at the above histogram of expenditure above it shows that expenditure has a positively skewed distribution.

I investigated the relationships between the other variable and expenditure dataset using the plots above.
We can see that there is a positive association between household income and expenditure in the scatter plot.
For the factor variables I used box plots to examine any differences present between the different levels within the factor variable and expenditure:
Different household tenure categories have different expenditure amounts with public rented households having a noticeably lower expenditure compared to the other 2 household levels
Households with a female head appear to have a lower expenditure than that of male household heads
Employment status appears to be associated with expenditure with unemployed having a dramatically lower expenditure compared to full and part time employment categories.
Household size appears to be positively correlated with expenditure, with larger households having higher expenditure except for 5+ person households which goes against this trend (further investigation?)
Number of adults in the household also appears positively correlated with households containing more adults having a higher expenditure.

@@

```{r,echo=F}
#Question 2
lm.income<-lm(expenditure~income,data = expend.df)
#summary(lm.income)
as.data.frame(tidy(lm.income)[,1:3])%>%
  kable(escape = F)%>%
  kable_styling(font_size = 15)
par(mfrow=c(2,2))
plot(lm.income)
```
Fitting a linear model $y_{i}=\beta_{0}+\beta_{1}x_{i}+\epsilon$ with the response $y$ the expenditure,the explanatory variable $x$ income and $i \in \{1,...,1200\}$ the ith household. The above table shows the estimated values of the intercept ($\beta_0$) and the estimated 'effect' of a 1 pound increase in income on expenditure $\beta_1$ and there standard errors.

@@

#### Part 5:
```{r,echo=F}
lm.log.income.sqd <- lm(log(expenditure)~income+I(income^2),data = expend.df)
tidy(lm.log.income.sqd)[,1:3]%>%
  kable(escape = F)%>%
  kable_styling(font_size = 15)
par(mfrow=c(2,2))
plot(lm.log.income.sqd)
```
Fitting the model $log(y_{i})=\beta_{0}+\beta_{1}x_{i}+\beta_{2}x^2_{i}+\epsilon$ gives the estimated coefficients above. Viewing the diagnostic plots the fitted vs residuals plot shows a fairly even scattering of points around 0 thus the linearity assumption appears valid. The QQ plot shows the points stay close to the straight line meaning the normality assumption is valid.
(scale location is slightly dogy- sign constant variance assumption has been violated)

@@

Using AIC as a measure of fit the best is model 4 as it has the lowest AIC value. Furthermore, model 4 has the highest $R^2$ and adjusted $R^2$ suggestion that model 4 'fits' the data best. This is further backed up by the diagnostic plots which indicate that this is the only model in which the linearity and normality assumptions are met. Due to all these reasons model 4 as my preferred model.

```{r}
tidy(lm.log.income.sqd)

mean(expend.df$income)
```
Using the model a 1 unit increase in household income increase the expenditure by $exp(\beta_1+2x\beta_2+\beta_2)-1$x100% of the old value. Subbing in the values of $\beta_1$ and $\beta_2$ this gives an increase by $exp(0.0032-0.0000029x-0.0000015)-1$x100%. Therefore for a household earning the mean income of £512 a £1 increase in income increases expenditure by approximately 0.17%. Using a more likely realistic increase in weekly income of £20 for a household earning £512 gives an increase of $exp(0.016-0.000015*20-0.000036)-1$x100% = 1.6%.

Include CI?

#### Task 7

@@

#### Model selection
For this model selection process I decided to only consider 2 way interactions as higher order interactions make interpretation of the model much more complicated.
When considering which interactions to include I first created plots to examine if there are any clear interactions between income and the other covariates. Considering the factor covarites thinking about the possible interactions none made much logical sense e.g. a household female head would have a different effect on expenditure to a male head depending on the number of people in the household. Thus didn't include any factor:factor interactions in the model.

```{r}
#interaction between sex and income
interaction.continous <-function(var){
  ggplot(data=expend.df,aes(x=income,y=expenditure,colour=.data[[var]]))+
    geom_point(size=0.1)
}

@@

```{r}
lm.log.full<-lm(log(expenditure)~income+I(income^2)+income*sex.hh+income*hh.adults+income*house.ten+lab.force*income+hh.size*income,data=expend.df)
step.method<- step(lm.log.full,trace=F)
tidy(step.method)
AIC(step.method)
glance(step.method)$adj.r.squared
plot(step.method)
#got rid of 2 interactions: income:sex and income:hhsize
```
